Predator clusters

### 2023-10-25 reports/01_predators.qmd
Clean up dist maps, add cluster env associations

library(sf)
library(targets)
library(tidyverse)

@@

The five most abundant species (in ind nmi\^-1).

@@

top_five <- predators_agg %>% 
  group_by(species) %>% 
  summarize(total_count = sum(count)) %>% 
  arrange(desc(total_count)) %>%
  slice(1:5)

top_five
```

@@

top_five_plot_data <- predators_agg %>% 
  semi_join(top_five, by = "species") %>% 
  group_by(year, species) %>% 
  summarize(total = sum(count), .groups = "drop") %>% 
  mutate(species = fct_reorder(species, total, .fun = sum),
         year = factor(year, levels = 2016:2012))
year_labels <- top_five_plot_data %>% 
  filter(species == "SNPT") %>% 
  arrange(desc(year)) %>%                   # Make the labels show up at
  mutate(total = cumsum(total) - total / 2) # correct y-axis position

ggplot(top_five_plot_data, aes(species, total, group = year)) +
  geom_col(color = "black", fill = "grey80") +
  geom_text(aes(label = year), year_labels) +
  labs(y = "Total species count") +
  theme_classic() +
  theme(axis.title.x = element_blank())
```

@@

library(terra)
# Mkae "top five" predator sightings spatially aware
topfive_sf <- predators_agg %>% 
  semi_join(top_five, by = "species") %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))
# Create a template raster of the study area
effort_ext <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  ext()
rast_template <- rast(crs = as.character(ant_proj())[[1]],
                      extent = effort_ext,
                      resolution = 30e3)
# Rasterize effort by year
effort_rast <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  rasterize(rast_template,
            field = "nmi",
            by = "year",
            fun = sum)
# Rasterize the predator sightings by year and species
topfive_rast <- topfive_sf %>% 
  st_transform(ant_proj()) %>% 
  mutate(year_species = interaction(year, species)) %>% 
  rasterize(rast_template, 
            field = "count",
            by = "year_species",
            fun = sum)
# Normalize counts by effort
topfive_norm <- topfive_rast
for (i in seq(nlyr(topfive_rast))) {
  lyr_name <- names(topfive_rast)[i]
  lyr_year <- substr(lyr_name, 1, 4)
  effort_year <- effort_rast[lyr_year] %>% 
    subst(0, NA)
  topfive_norm[[i]] <- topfive_norm[[i]] / effort_year
}
# Average across years
topfive_climatology <- unique(top_five$species) %>% 
  map(function(sp) {
    sp_rast <- subset(topfive_norm, 
                      which(str_detect(names(topfive_norm), sp)))
    mean(sp_rast, na.rm = TRUE)
  }) %>% 
  set_names(unique(top_five$species)) %>% 
  rast()
# Classify raster (otherwise CRSE throw it all off)
topfive_clim_class <- topfive_climatology %>% 
  classify(rcl = c(0, 1, 5, 10, 50))

@@

# Make the map! 
ant_basemap() +
  tidyterra::geom_spatraster(data = topfive_clim_class) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_wrap(~lyr) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "top")

@@

Distribution maps of the five most abundant species in each survey year, aggregated to 30km bins.

```{r}
#| label: topfive_map
#| warning: false
#| column: screen
#| out-width: 100%
#| fig-width: 6
#| fig-dpi: 600

@@

topfive_class <- topfive_rast %>% 
  classify(rcl = c(0, 1, 5, 10, 50, 100, 1000, Inf))

topfive_class_df <- as.data.frame(topfive_class, xy = TRUE) %>% 
  pivot_longer(-c("x", "y"), 
               names_to = "name", 
               values_to = "count") %>% 
  separate_wider_delim(name, 
                       delim = ".", 
                       names = c("year", "species")) %>% 
  drop_na(count)

ant_basemap() +
  geom_raster(aes(x, y, fill = count), topfive_class_df) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_grid(species ~ year) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "right",
        text = element_text(size = 8))
```

Distribution maps of five species of interest in each survey year, also aggregated to 25km bins.

@@

>>>>

library(sf)
library(targets)
library(terra)
library(tidyverse)

@@

The six most abundant species (in ind nmi\^-1).

@@

top_six <- predators_agg %>% 
  group_by(species) %>% 
  summarize(total_count = sum(count)) %>% 
  arrange(desc(total_count)) %>%
  slice(1:6) %>% 
  mutate(common_name = c("Snow petrel", 
                         "Antarctic fur seal",
                         "Adélie penguin",
                         "Crabeater seal",
                         "Antarctic petrel",
                         "Southern fulmar"))

top_six
```

@@

top_six_plot_data <- predators_agg %>% 
  semi_join(top_six, by = "species") %>% 
  group_by(year, species) %>% 
  summarize(total = sum(count), .groups = "drop") %>% 
  mutate(species = fct_reorder(species, total, .fun = sum),
         year = factor(year, levels = 2016:2012))
year_labels <- top_six_plot_data %>% 
  filter(species == "SNPT") %>% 
  arrange(desc(year)) %>%                   # Make the labels show up at
  mutate(total = cumsum(total) - total / 2) # correct y-axis position

ggplot(top_six_plot_data, aes(species, total, group = year)) +
  geom_col(color = "black", fill = "grey80") +
  geom_text(aes(label = year), year_labels) +
  labs(y = "Total species count") +
  theme_classic() +
  theme(axis.title.x = element_blank())
```

@@

# Make "top six" predator sightings spatially aware
topsix_sf <- predators_agg %>% 
  semi_join(top_six, by = "species") %>% 
  left_join(select(top_six, species, common_name), by = "species") %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))
# Create a template raster of the study area
effort_ext <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  ext()
rast_template <- rast(crs = as.character(ant_proj())[[1]],
                      extent = effort_ext,
                      resolution = 30e3)
# Rasterize effort by year
effort_rast <- effort_sf %>% 
  st_transform(ant_proj()) %>% 
  rasterize(rast_template,
            field = "nmi",
            by = "year",
            fun = sum)
# Rasterize the predator sightings by year and species
topsix_rast <- topsix_sf %>% 
  st_transform(ant_proj()) %>% 
  mutate(year_commonname = interaction(year, common_name)) %>% 
  rasterize(rast_template, 
            field = "count",
            by = "year_commonname",
            fun = sum)
# Normalize counts by effort
topsix_norm <- topsix_rast
for (i in seq(nlyr(topsix_rast))) {
  lyr_name <- names(topsix_rast)[i]
  lyr_year <- substr(lyr_name, 1, 4)
  effort_year <- effort_rast[lyr_year] %>% 
    subst(0, NA)
  topsix_norm[[i]] <- topsix_norm[[i]] / effort_year
}
# Average across years
topsix_climatology <- unique(top_six$common_name) %>% 
  map(function(sp) {
    sp_rast <- subset(topsix_norm, 
                      which(str_detect(names(topsix_norm), sp)))
    mean(sp_rast, na.rm = TRUE)
  }) %>% 
  set_names(unique(top_six$common_name)) %>% 
  rast()
# Classify raster (otherwise CRSE throw it all off)
topsix_clim_class <- topsix_climatology %>% 
  classify(rcl = c(0, 1, 5, 10, 50))

@@

# Make the map! 
ant_basemap() +
  tidyterra::geom_spatraster(data = topsix_clim_class) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_wrap(~lyr) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "top")

@@

Distribution maps of the six most abundant species in each survey year, aggregated to 30km bins.

```{r}
#| label: topsix_map
#| warning: false
#| column: screen
#| out-width: 100%
#| fig-width: 6
#| fig-dpi: 600

@@

topsix_class <- topsix_rast %>% 
  classify(rcl = c(0, 1, 5, 10, 50, 100, 1000, Inf))

topsix_class_df <- as.data.frame(topsix_class, xy = TRUE) %>% 
  pivot_longer(-c("x", "y"), 
               names_to = "name", 
               values_to = "count") %>% 
  separate_wider_delim(name, 
                       delim = ".", 
                       names = c("year", "species")) %>% 
  drop_na(count)

ant_basemap() +
  geom_raster(aes(x, y, fill = count), topsix_class_df) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_grid(species ~ year) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "right",
        text = element_text(size = 8))
```

Distribution maps of six species of interest in each survey year, also aggregated to 25km bins.

@@

Distribution maps of all species in all years, for supplemental.

```{r}
#| label: pred_map
#| warning: false
#| column: screen
#| out-width: 100%
#| fig-height: 9
#| fig-width: 7.5
#| fig-dpi: 600

pred_sf <- predators_agg %>% 
  filter(!str_starts(species, "U")) %>% 
  mutate(common_name = code_to_common(species)) %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))

pred_rast <- pred_sf %>% 
  st_transform(ant_proj()) %>% 
  mutate(year_commonname = interaction(year, common_name)) %>% 
  rasterize(rast_template, 
            field = "count",
            by = "year_commonname",
            fun = sum)

# Normalize counts by effort
pred_norm <- pred_rast
for (i in seq(nlyr(pred_rast))) {
  lyr_name <- names(pred_rast)[i]
  lyr_year <- substr(lyr_name, 1, 4)
  effort_year <- effort_rast[lyr_year] %>% 
    subst(0, NA)
  pred_norm[[i]] <- pred_norm[[i]] / effort_year
}
# Average across years
pred_climatology <- unique(code_to_common(pred_sf$species)) %>% 
  map(function(sp) {
    sp_rast <- subset(pred_norm, 
                      which(str_detect(names(pred_norm), sp)))
    mean(sp_rast, na.rm = TRUE)
  }) %>% 
  set_names(unique(pred_sf$common_name)) %>% 
  rast()

pred_class <- pred_climatology %>% 
  classify(rcl = c(0, 1, 5, 10, 50, 100, 1000))

pred_class_df <- as.data.frame(pred_class, xy = TRUE) %>% 
  pivot_longer(-c("x", "y"), 
               names_to = "name", 
               values_to = "count") %>% 
  drop_na(count) %>% 
  mutate(name = map_chr(name, \(n) paste(strwrap(n, 15), collapse = "\n")))

p <- ant_basemap() +
  geom_raster(aes(x, y, fill = count), pred_class_df) +
  geom_sf(data = tar_read("ant_sf", store = here::here("_targets"))) +
  scale_fill_brewer(palette = "RdPu", 
                    na.value = NA, 
                    direction = 1,
                    na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-64, -54, by = 4)) +
  scale_y_continuous(breaks = seq(-64, -59, by = 2)) +
  labs(fill = "Ind nmi^-1") +
  facet_wrap(~ name, ncol = 6) +
  coord_sf(xlim = map_bbox[1:2],
           ylim = map_bbox[3:4],
           crs = ant_proj()) +
  theme(legend.position = "top",
        text = element_text(size = 8))

ggsave("figs/s1.jpg", p, width = 6, height = 8, units = "in", dpi = 600)
```

### 2023-10-25 reports/04_pred_clus.qmd
Clean up dist maps, add cluster env associations

-   Cluster 1 = pack ice
-   Cluster 2 = marginal ice
-   Cluster 3 = open water

@@

# Indicator species grob
indval_grob <- apply(sightings_indval$indval, 2, function(col) {
  # Extract indicator values >0.25 from each cluster's column and format with
  # species name
  col <- sort(col[col > 0.25], decreasing = TRUE)
  species <- code_to_common(names(col))
  if (length(col) == 0) {
    "None"
  } else {
    str_glue("{species} ({round(col, 2)})")
  }
}) %>% 
  # Create a 
  map(\(x) {
    nrow = max(map_dbl(., length))
    c(x, rep("", nrow - length(x)))
  }) %>% 
  as_tibble() %>% 
  set_names(levels(stations_clust$pred_clust)) %>% 
  tableGrob(
    rows = NULL,
    theme = ttheme_minimal(base_size = 6,
                           padding = unit(c(1, 1), "mm"),
                           core = list(fg_params = list(hjust = 0,
                                                        x = 0)),
                           colhead = list(fg_params = list(hjust = 0,
                                                           x = 0)))
  )

ant_basemap(map_lim) +
  # Sea ice
  geom_tile(aes(x, y, fill = seaice_conc), 
            filter(seaice_conc_df, between(seaice_conc, 0.65, 1)),
            alpha = 0.8) +
  scale_fill_distiller(lim = c(0.65, 1),
                       palette = "Blues",
                       na.value = "transparent") +
  # geom_sf(data = seaice_contours, color = "darkblue") +
  # Predator clusters
  ggnewscale::new_scale_fill() +
  geom_sf(aes(fill = pred_clust), 
          stations_clust, 
          size = 2, 
          shape = 22) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(override.aes = list(size = 3))) +
  coord_ant(map_lim) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  inset_element(indval_grob, 
                0.67, 0, 1, 0.5, 
                align_to = "panel")
```

@@

@@

Median ice coverage increased from the open water cluster (median 33.0%) to the marginal ice (59.6%) and pack ice (70.6%) clusters. This supports the labels of "open water" and "marginal ice" for clusters 3 and 2, respectively. Also provides evidence for the "pack ice" cluster.

>>>>

-   Cluster 1 = pack ice
-   Cluster 2 = open water
-   Cluster 3 = marginal ice

@@

ant_basemap(map_lim) +
  # Sea ice
  geom_tile(aes(x, y, fill = seaice_conc), 
            filter(seaice_conc_df, between(seaice_conc, 0.65, 1)),
            alpha = 0.8) +
  scale_fill_distiller(lim = c(0.65, 1),
                       palette = "Blues",
                       na.value = "transparent") +
  # Predator clusters
  ggnewscale::new_scale_fill() +
  geom_sf(aes(fill = pred_clust), 
          stations_clust, 
          size = 2, 
          shape = 22) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(override.aes = list(size = 3))) +
  coord_ant(map_lim) +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

```{r}
#| label: cluster_maps2
#| warning: false

library(eks)
cluster_kde <- stations_clust %>% 
  group_by(pred_clust) %>% 
  group_map(\(rows, keys) st_kde(rows)$sf %>% 
              filter(contlabel %in% c(50, 95)) %>% 
              mutate(pred_clust = keys$pred_clust)) %>% 
  reduce(sf:::rbind.sf)

ant_basemap(map_lim) +
  # Predator clusters
  ggnewscale::new_scale_fill() +
  geom_sf(aes(color = pred_clust,
              linetype = contlabel), 
          cluster_kde, 
          fill = NA,
          linewidth = 1) +
  facet_wrap(~ pred_clust) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_color_brewer(palette = "Dark2") +
  scale_linetype_manual(values = c(2, 1), guide = "none") +
  coord_ant(map_lim) +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

@@

## Environmental conditions by cluster

```{r}
#| label: env_clust

kw <- map(c("zuml_m", "avg.temp", "avg.salinity", 
            "Integ.chla.100m", "Integ.phae.100m"),
          \(v) {
            kw <- kruskal.test(stations_clust[[v]], stations_clust$pred_clust)
            data.frame(var = v, p = kw$p.value * 5)
          }) %>% 
  list_rbind()

stations_clust %>% 
  as.data.frame() %>% 
  group_by(pred_clust) %>%
  summarize(
    across(c(zuml_m, avg.temp, avg.salinity, Integ.chla.100m, Integ.phae.100m),
           list(mean = partial(mean, na.rm = TRUE), 
                q1 = \(x) quantile(x, 0.25, na.rm = TRUE), 
                q3 = \(x) quantile(x, 0.75, na.rm = TRUE)),
           .names = "{.col}={.fn}")
  ) %>% 
  pivot_longer(-pred_clust) %>% 
  separate_wider_delim(name, "=", names = c("var", "fn")) %>% 
  pivot_wider(names_from = "fn", values_from = "value") %>% 
  mutate(across(mean:q3, \(x) formatC(signif(x, digits=3), digits = 3, format = "fg", flag = "#"))) %>% 
  transmute(pred_clust, var, value = str_glue("{mean} ({q1} - {q3})")) %>% 
  pivot_wider(names_from = pred_clust, values_from = "value") %>% 
  left_join(kw, by = "var") %>% 
  mutate(var = ifelse(p <= 0.05, paste0(var, " *"), var)) %>% 
  select(-p) %>% 
  knitr::kable()

@@

Median ice coverage increased from the open water cluster (median 34.1%) to the marginal ice (56.0%) and pack ice (74.9%) clusters. This supports the labels of "open water" and "marginal ice" for clusters 3 and 2, respectively. Also provides evidence for the "pack ice" cluster.
